---
title: "Bayesian Analysis of Stable Isotope Variation at La Ventilla, Teotihuacan"
author: "Gina M. Buckley"
date: "2026-03-10"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE
)
```

# Introduction

This document provides a reproducible workflow describing the Bayesian statistical analyses used to evaluate stable isotope variation for paleodietary reconstruction at the La Ventilla district of Teotihuacan. The code implements Bayesian models to estimate differences in isotopic values across social, biological, and temporal groups within the skeletal assemblage.

To replicate the analyses presented here, users must first install the R package **brms**, which provides an interface to the Stan probabilistic programming language for Bayesian inference.

### Load required packages

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(brms)
library(tidybayes)
library(broom.mixed)
library(writexl)
library(bayesplot)
library(ggridges)
library(posterior)
library(patchwork)

set.seed(123)
```

### Load isotopic dataset
Stable isotope data from individuals recovered at the Compound of the Glyphs (LV2) and the Compound of the Artisans (LV3) were loaded into R for analysis.

```{r}
df_ventilla_isotopes <- read_csv("https://raw.githubusercontent.com/gbuck04/la-ventilla-diet/refs/heads/main/LV_data.csv")
df_teo_compounds <- read_csv("https://raw.githubusercontent.com/gbuck04/la-ventilla-diet/refs/heads/main/Teo_neighborhood_isotope_data.csv")
```

The dataset contains two extreme δ¹³C_collagen values (−20.3‰ in LV2 and −18.1‰ in LV3). Rather than removing these observations, they were retained in the analysis. Bayesian models estimate parameters through posterior distributions, which incorporate uncertainty directly and are generally less sensitive to deviations from normality and the influence of individual observations compared with traditional NHST approaches. Retaining these values therefore allows the full observed variation in the dataset to be represented in the model.


### Data cleaning and preparation

The dataset was screened for preservation quality prior to analysis. Samples with carbon-to-phosphorus (C:I) ratios outside the accepted range for well-preserved bioapatite were excluded from apatite-based isotopic analyses. Specifically, δ¹³C_apatite and δ¹⁸O_apatite values were set to missing (NA) when the C:I ratio fell outside the commonly accepted range of 2.8–4.0, indicating potentially altered or poorly preserved apatite.

```{r}
df_clean_isotopes <- df_ventilla_isotopes %>%
mutate(
  c_ap = ifelse(CI < 2.8 | CI > 4.0, NA, c_ap),
  o_ap = ifelse(CI < 2.8 | CI > 4.0, NA, o_ap)
)
```


### Filter published isotopic data to generate priors

To parameterize informative priors for the Bayesian models, summary statistics were calculated from previously published stable isotope data from Teotihuacan residential compounds. Published values for δ¹³C (collagen), δ¹⁵N (collagen), δ¹³C (apatite), and δ¹⁸O (apatite) were compiled into a reference dataset of isotopic measurements from multiple compounds within the city.

Samples originating from La Ventilla, Frente 2 and Frente 3 were excluded from this dataset to ensure that the priors were derived from independent data. In addition, samples from outside of Teotihuacan were removed. Because the new data presented in this study derive from bone collagen and bone bioapatite, measurements from other components (e.g., enamel bioapatite) were also excluded. The resulting dataset therefore represents isotopic variation from previously published Teotihuacan skeletal samples that are comparable to the materials analyzed in this study.


```{r}
df_teo_filtered <- df_teo_compounds %>%
  filter(
    !provenance %in% c("La Ventilla, Frente 2", "La Ventilla, Frente 3", "Copan", "Kaminaljuyu", "Monte Alban", "Copan Valley"),
    component != "enamel bioapatite"
  )
```

### Generate priors from published data

Previous NHST results indicate that mean isotopic values do not differ significantly among residential compounds within Teotihuacan. Informative priors were used to stabilize parameter estimation given modest sample sizes while incorporating well-established isotopic baselines for Teotihuacan populations. Therefore, the filtered dataset was combined to calculate overall summary statistics for δ¹³C collagen, δ¹⁵N collagen, δ¹³C apatite, and δ¹⁸O apatite. The mean and standard deviation of each isotopic system were used to parameterize normal prior distributions for the intercept terms in the Bayesian models.


```{r}
generate_prior <- function(data, variable, class = "Intercept") {
  stats <- data %>%
    filter(!is.na(.data[[variable]])) %>%
    summarise(mean_val = mean(.data[[variable]]), sd_val = sd(.data[[variable]]))
  sprintf("prior(normal(%.2f, %.2f), class = '%s')", stats$mean_val, stats$sd_val, class)
}

prior_strings <- sapply(c("c_col", "n_air", "c_ap", "o_ap"), function(var) generate_prior(df_teo_filtered, var))
cat(paste0(prior_strings, collapse = "\n"))
```

### Bayesian modeling framework

Bayesian regression models were implemented using the **brms** package. Each isotopic system was modeled separately, with group membership included as a fixed effect. Posterior distributions were estimated using four Markov Chain Monte Carlo chains with 4000 iterations per chain (1000 warmup), providing 12,000 post-warmup posterior draws for inference.
Model convergence was assessed using R̂ statistics and effective sample sizes for all parameters. All models showed satisfactory convergence (R̂ values approximately 1.00).


---


## Residential compound comparison (LV2 vs LV3)

### Subset data
```{r}
df_lv_subset <- df_clean_isotopes %>%
filter(frente %in% c("LV2", "LV3")) %>%
mutate(frente = factor(frente))

table(df_lv_subset$frente)
```

### Define priors
```{r}
common_priors <- set_prior("student_t(3, 0, 2.5)", class = "sigma")
group_prior <- set_prior("normal(0, 2)", class = "b")
```

### Fit Bayesian model
```{r}

# d13C collagen
model_c_col <- brm(
  c_col ~ frente, 
  data = df_lv_subset, 
  prior = c(
    set_prior("normal(-8.21, 0.95)", class = "Intercept"),
    group_prior, 
    common_priors), 
  iter = 4000, warmup = 1000, chains = 4, cores = 4, seed = 123)

# d15N
model_n_air <- brm(
  n_air ~ frente, 
  data = df_lv_subset, 
  prior = c(
    set_prior("normal(9.22, 1.02)", class = "Intercept"),
    group_prior, 
    common_priors), 
  iter = 4000, warmup = 1000, chains = 4, cores = 4, seed = 124)

# d13C apatite
model_c_ap <- brm(
  c_ap ~ frente, 
  data = df_lv_subset, 
  prior = c(
    set_prior("normal(-3.17, 0.69)", class = "Intercept"),
    group_prior, 
    common_priors), 
   iter = 4000, warmup = 1000, chains = 4, cores = 4, seed = 125)

# d18O apatite
model_o_ap <- brm(
  o_ap ~ frente, 
  data = df_lv_subset, 
  prior = c(
    set_prior("normal(-8.39, 1.32)", class = "Intercept"),
    group_prior, 
    common_priors), 
iter = 4000, warmup = 1000, chains = 4, cores = 4, seed = 126)
```


### Posterior summaries
```{r}
# Tidy fixed effects
summaries <- list(
  tidy(model_c_col, effects = "fixed", conf.int = TRUE) %>% mutate(model = "δ13C (collagen)"),
  tidy(model_n_air, effects = "fixed", conf.int = TRUE) %>% mutate(model = "δ15N (collagen)"),
  tidy(model_c_ap, effects = "fixed", conf.int = TRUE) %>% mutate(model = "δ13C (apatite)"),
  tidy(model_o_ap, effects = "fixed", conf.int = TRUE) %>% mutate(model = "δ18O (apatite)")
)

posterior_table <- bind_rows(summaries) %>%
  select(model, term, estimate, std.error, conf.low, conf.high) %>%
  rename(
    Isotope_System = model,
    Parameter = term,
    Estimate = estimate,
    Std_Error = std.error,
    CI_2.5 = conf.low,
    CI_97.5 = conf.high
)

print(posterior_table)
```

### Posterior probabilities: P(LV2 > LV3)
```{r}
# δ13C (collagen)
post_c_col <- as_draws_df(model_c_col)
mean(post_c_col$b_frenteLV3 < 0, na.rm = TRUE) # P(LV2 > LV3)
mcmc_areas(model_c_col, pars = c("b_Intercept", "b_frenteLV3"))


# δ15N
post_n_air <- as_draws_df(model_n_air)
mean(post_n_air$b_frenteLV3 < 0, na.rm = TRUE)
mcmc_areas(model_n_air, pars = c("b_Intercept", "b_frenteLV3"))


# δ13C (apatite)
post_c_ap <- as_draws_df(model_c_ap)
mean(post_c_ap$b_frenteLV3 < 0, na.rm = TRUE)
mcmc_areas(model_c_ap, pars = c("b_Intercept", "b_frenteLV3"))


# δ18O
post_o_ap <- as_draws_df(model_o_ap)
mean(post_o_ap$b_frenteLV3 < 0, na.rm = TRUE)
mcmc_areas(model_o_ap, pars = c("b_Intercept", "b_frenteLV3"))
```


```{r}
# Extract posterior samples
post_lv_c_col <- as_draws_df(model_c_col)
post_lv_n_air <- as_draws_df(model_n_air)
post_lv_c_ap  <- as_draws_df(model_c_ap)
post_lv_o_ap  <- as_draws_df(model_o_ap)

# Calculate posterior probabilities
prob_lv2_gt_lv3 <- tibble(
  Isotope_System = c("δ13C (collagen)", "δ15N (collagen)", "δ13C (apatite)", "δ18O (apatite)"),
  P_LV2_gt_LV3 = c(
    mean(post_lv_c_col$b_frenteLV3 < 0, na.rm = TRUE),  # LV2 > LV3 for δ13C_collagen
    mean(post_lv_n_air$b_frenteLV3 < 0, na.rm = TRUE),
    mean(post_lv_c_ap$b_frenteLV3 < 0, na.rm = TRUE),
    mean(post_lv_o_ap$b_frenteLV3 < 0, na.rm = TRUE)
  )
)

print(prob_lv2_gt_lv3)
```

Models with heteroskedastic variance: sigma ~ frente

```{r}
# d13Ccol
model_c_col_sd <- brm(
  bf(c_col ~ frente, sigma ~ frente),
  data = df_lv_subset,
  prior = c(
    set_prior("normal(-8.21, 0.95)", class = "Intercept"),
    set_prior("normal(0, 2)", class = "b"),
    set_prior("student_t(3, 0, 2.5)", class = "Intercept", dpar = "sigma"),
    set_prior("normal(0, 1)", class = "b", dpar = "sigma")
  ),
  iter = 4000, warmup = 1000, chains = 4, cores = 4, seed = 123,
  control = list(adapt_delta = 0.95, max_treedepth = 12)
)

# δ15N (collagen)
model_n_air_sd <- brm(
  bf(n_air ~ frente, sigma ~ frente),
  data = df_lv_subset,
  prior = c(
    set_prior("normal(9.22, 1.02)", class = "Intercept"),
    set_prior("normal(0, 2)", class = "b"),
    set_prior("student_t(3, 0, 2.5)", class = "Intercept", dpar = "sigma"),
    set_prior("normal(0, 1)", class = "b", dpar = "sigma")
  ),
  iter = 4000, warmup = 1000, chains = 4, cores = 4, seed = 123,
  control = list(adapt_delta = 0.95, max_treedepth = 12)
)

# δ13C (apatite)
model_c_ap_sd <- brm(
  bf(c_ap ~ frente, sigma ~ frente),
  data = df_lv_subset,
  prior = c(
    set_prior("normal(-3.17, 0.69)", class = "Intercept"),
    set_prior("normal(0, 2)", class = "b"),
    set_prior("student_t(3, 0, 2.5)", class = "Intercept", dpar = "sigma"),
    set_prior("normal(0, 1)", class = "b", dpar = "sigma")
  ),
  iter = 4000, warmup = 1000, chains = 4, cores = 4, seed = 123,
  control = list(adapt_delta = 0.95, max_treedepth = 12)
)

# δ18O (apatite)
model_o_ap_sd <- brm(
  bf(o_ap ~ frente, sigma ~ frente),
  data = df_lv_subset,
  prior = c(
    set_prior("normal(-8.39, 1.32)", class = "Intercept"),
    set_prior("normal(0, 2)", class = "b"),
    set_prior("student_t(3, 0, 2.5)", class = "Intercept", dpar = "sigma"),
    set_prior("normal(0, 1)", class = "b", dpar = "sigma")
  ),
  iter = 4000, warmup = 1000, chains = 4, cores = 4, seed = 123,
  control = list(adapt_delta = 0.95, max_treedepth = 12)
)

```

### Posterior variance estimates: P(LV2 more variable than LV3)

```{r} 
# Extract posterior probabilities that LV2 is more variable than LV3
get_sigma_prob <- function(model) {
  posterior <- as_draws_df(model)
  mean(posterior$b_sigma_frenteLV3 < 0, na.rm = TRUE)  # P(LV2 > LV3)
}

get_sigma_stats <- function(model) {
  posterior <- as_draws_df(model)
  diff <- posterior$b_sigma_frenteLV3

  tibble(
    P_LV2_more_variable = mean(diff < 0, na.rm = TRUE),
    Posterior_Mean = mean(diff, na.rm = TRUE),
    Posterior_SD = sd(diff, na.rm = TRUE),
    CI_2.5 = quantile(diff, 0.025, na.rm = TRUE),
    CI_97.5 = quantile(diff, 0.975, na.rm = TRUE)
  )
}

prob_variance_lv2_gt_lv3 <- tibble(
  Isotope_System = c("δ13C (collagen)", "δ15N (collagen)", "δ13C (apatite)", "δ18O (apatite)"),
  P_LV2_more_variable = c(
    get_sigma_prob(model_c_col_sd),
    get_sigma_prob(model_n_air_sd),
    get_sigma_prob(model_c_ap_sd),
    get_sigma_prob(model_o_ap_sd)
  )
)

print(prob_variance_lv2_gt_lv3)
```

### Visualize data

```{r}
ggplot(df_lv_subset, aes(x = c_col, y = frente, fill = frente)) +
  geom_density_ridges(alpha = 0.6, scale = 1, na.rm = TRUE) +
  labs(title = "δ13C (collagen) by Compound", x = "δ13C (‰)", y = "Compound") +
  theme_minimal() +
  theme(legend.position = "none")

ggplot(df_lv_subset, aes(x = n_air, y = frente, fill = frente)) +
  geom_density_ridges(alpha = 0.6, scale = 1, na.rm = TRUE) +
  labs(title = "δ15N (collagen) by Compound", x = "δ15N (‰)", y = "Compound") +
  theme_minimal() +
  theme(legend.position = "none")

ggplot(df_lv_subset, aes(x = c_ap, y = frente, fill = frente)) +
  geom_density_ridges(alpha = 0.6, scale = 1, na.rm = TRUE) +
  labs(title = "δ13C (apatite) by Compound", x = "δ13C (‰)", y = "Compound") +
  theme_minimal() +
  theme(legend.position = "none")

ggplot(df_lv_subset, aes(x = o_ap, y = frente, fill = frente)) +
  geom_density_ridges(alpha = 0.6, scale = 1, na.rm = TRUE) +
  labs(title = "δ18O (apatite) by Compound", x = "δ18O (‰)", y = "Compound") +
  theme_minimal() +
  theme(legend.position = "none")
```

---

## Biological Sex Comparisons

### Subset data
```{r}
df_sex_subset <- df_clean_isotopes %>%
  filter(sex %in% c("M", "F")) %>%
  mutate(sex = factor(sex, levels = c("F", "M")))

table(df_sex_subset$sex)
```

### Fit Bayesian model
```{r}

# d13C collagen
model_c_col_sex <- brm(
  c_col ~ sex, 
  data = df_sex_subset, 
  prior = c(
    set_prior("normal(-8.21, 0.95)", class = "Intercept"),
    group_prior, 
    common_priors), 
  iter = 4000, warmup = 1000, chains = 4, cores = 4, seed = 131)

# d15N
model_n_air_sex <- brm(
  n_air ~ sex, 
  data = df_sex_subset, 
  prior = c(
    set_prior("normal(9.22, 1.02)", class = "Intercept"),
    group_prior, 
    common_priors), 
  iter = 4000, warmup = 1000, chains = 4, cores = 4, seed = 132)

# d13C apatite
model_c_ap_sex <- brm(
  c_ap ~ sex, 
  data = df_sex_subset, 
  prior = c(
    set_prior("normal(-3.17, 0.69)", class = "Intercept"),
    group_prior, 
    common_priors), 
  iter = 4000, warmup = 1000, chains = 4, cores = 4, seed = 133)

# d18O apatite
model_o_ap_sex <- brm(
  o_ap ~ sex, 
  data = df_sex_subset, 
  prior = c(
    set_prior("normal(-8.39, 1.32)", class = "Intercept"),
    group_prior, 
    common_priors), 
iter = 4000, warmup = 1000, chains = 4, cores = 4, seed = 134)
```

### Posterior summaries

```{r}
# Tidy fixed effects
summaries <- list(
  tidy(model_c_col_sex, effects = "fixed", conf.int = TRUE) %>% mutate(model = "δ13C (collagen)"),
  tidy(model_n_air_sex, effects = "fixed", conf.int = TRUE) %>% mutate(model = "δ15N (collagen)"),
  tidy(model_c_ap_sex, effects = "fixed", conf.int = TRUE) %>% mutate(model = "δ13C (apatite)"),
  tidy(model_o_ap_sex, effects = "fixed", conf.int = TRUE) %>% mutate(model = "δ18O (apatite)")
)

posterior_table_sex <- bind_rows(summaries) %>%
  select(model, term, estimate, std.error, conf.low, conf.high) %>%
  rename(
    Isotope_System = model,
    Parameter = term,
    Estimate = estimate,
    Std_Error = std.error,
    CI_2.5 = conf.low,
    CI_97.5 = conf.high
)

print(posterior_table_sex)
```

### Posterior Probabilities: P(M > F)

```{r}
post_c_col_sex <- as_draws_df(model_c_col_sex)
post_n_air_sex <- as_draws_df(model_n_air_sex)
post_c_ap_sex  <- as_draws_df(model_c_ap_sex)
post_o_ap_sex  <- as_draws_df(model_o_ap_sex)

post_sex_probs <- tibble(
  Isotope_System = c("δ13C (collagen)", "δ15N (collagen)", "δ13C (apatite)", "δ18O (apatite)"),
  P_M_gt_F = c(
    mean(post_c_col_sex$b_sexM > 0, na.rm = TRUE),
    mean(post_n_air_sex$b_sexM > 0, na.rm = TRUE),
    mean(post_c_ap_sex$b_sexM > 0, na.rm = TRUE),
    mean(post_o_ap_sex$b_sexM > 0, na.rm = TRUE)
  )
)

print(post_sex_probs)
```

Dietary Variability: sigma ~ sex

```{r}
model_c_col_sd_sex <- brm(
  bf(c_col ~ sex, sigma ~ sex),
  data = df_sex_subset,
  prior = c(
    set_prior("normal(-8.21, 0.95)", class = "Intercept"),
    set_prior("normal(0, 2)", class = "b"),
    set_prior("student_t(3, 0, 2.5)", class = "Intercept", dpar = "sigma"),
    set_prior("normal(0, 1)", class = "b", dpar = "sigma")
  ),
  iter = 4000, warmup = 1000,
  chains = 4, cores = 4, seed = 701,
  control = list(adapt_delta = 0.95, max_treedepth = 12)
)

model_n_air_sd_sex <- brm(
  bf(n_air ~ sex, sigma ~ sex),
  data = df_sex_subset,
  prior = c(
    set_prior("normal(9.22, 1.02)", class = "Intercept"),
    set_prior("normal(0, 2)", class = "b"),
    set_prior("student_t(3, 0, 2.5)", class = "Intercept", dpar = "sigma"),
    set_prior("normal(0, 1)", class = "b", dpar = "sigma")
  ),
  iter = 4000, warmup = 1000,
  chains = 4, cores = 4, seed = 702,
  control = list(adapt_delta = 0.95, max_treedepth = 12)
)

model_c_ap_sd_sex <- brm(
  bf(c_ap ~ sex, sigma ~ sex),
  data = df_sex_subset,
  prior = c(
    set_prior("normal(-3.17, 0.69)", class = "Intercept"),
    set_prior("normal(0, 2)", class = "b"),
    set_prior("student_t(3, 0, 2.5)", class = "Intercept", dpar = "sigma"),
    set_prior("normal(0, 1)", class = "b", dpar = "sigma")
  ),
  iter = 4000, warmup = 1000,
  chains = 4, cores = 4, seed = 703,
  control = list(adapt_delta = 0.95, max_treedepth = 12)
)

model_o_ap_sd_sex <- brm(
  bf(o_ap ~ sex, sigma ~ sex),
  data = df_sex_subset,
  prior = c(
    set_prior("normal(-8.39, 1.32)", class = "Intercept"),
    set_prior("normal(0, 2)", class = "b"),
    set_prior("student_t(3, 0, 2.5)", class = "Intercept", dpar = "sigma"),
    set_prior("normal(0, 1)", class = "b", dpar = "sigma")
  ),
  iter = 4000, warmup = 1000,
  chains = 4, cores = 4, seed = 704,
  control = list(adapt_delta = 0.95, max_treedepth = 12)
)

# Compare group variance: P(M > F)
get_sigma_prob_sex <- function(model) {
  posterior <- as_draws_df(model)
  mean(posterior$b_sigma_sexM > 0, na.rm = TRUE)  # P(M more variable)
}

get_sigma_stats_sex <- function(model) {
  posterior <- as_draws_df(model)
  diff <- posterior$b_sigma_sexM
  tibble(
    P_M_more_variable = mean(diff > 0, na.rm = TRUE),
    P_F_more_variable = mean(diff < 0, na.rm = TRUE),
    Posterior_Mean = mean(diff, na.rm = TRUE),
    Posterior_SD = sd(diff, na.rm = TRUE),
    CI_2.5 = quantile(diff, 0.025, na.rm = TRUE),
    CI_97.5 = quantile(diff, 0.975, na.rm = TRUE)
  )
}

prob_variance_sex <- tibble(
  Isotope_System = c("δ13C (collagen)", "δ15N (collagen)", "δ13C (apatite)", "δ18O (apatite)"),
  P_Male_more_variable = c(
    get_sigma_prob_sex(model_c_col_sd_sex),
    get_sigma_prob_sex(model_n_air_sd_sex),
    get_sigma_prob_sex(model_c_ap_sd_sex),
    get_sigma_prob_sex(model_o_ap_sd_sex)
  )
)

prob_variance_sex
```

### Visualize data

```{r}
ggplot(df_sex_subset, aes(x = c_col, y = sex, fill = sex)) +
  geom_density_ridges(alpha = 0.6, scale = 1, na.rm = TRUE) +
  labs(title = "δ13C (collagen) by Sex", x = "δ13C (‰)", y = "Sex") +
  theme_minimal() +
  theme(legend.position = "none")

ggplot(df_sex_subset, aes(x = n_air, y = sex, fill = sex)) +
  geom_density_ridges(alpha = 0.6, scale = 1, na.rm = TRUE) +
  labs(title = "δ15N (collagen) by Sex", x = "δ15N (‰)", y = "Sex") +
  theme_minimal() +
  theme(legend.position = "none")

ggplot(df_sex_subset, aes(x = c_ap, y = sex, fill = sex)) +
  geom_density_ridges(alpha = 0.6, scale = 1, na.rm = TRUE) +
  labs(title = "δ13C (apatite) by Sex", x = "δ13C (‰)", y = "Sex") +
  theme_minimal() +
  theme(legend.position = "none")

ggplot(df_sex_subset, aes(x = o_ap, y = sex, fill = sex)) +
  geom_density_ridges(alpha = 0.6, scale = 1, na.rm = TRUE) +
  labs(title = "δ18O (apatite) by Sex", x = "δ18O (‰)", y = "Sex") +
  theme_minimal() +
  theme(legend.position = "none")
```


---

## Geographic Origin (Local vs Nonlocal)
Individuals were classified as local or nonlocal based on δ¹⁸O apatite values using a previously defined local isotopic range.

## Subset data

```{r}
df_locality_subset <- df_clean_isotopes %>%
  filter(!is.na(o_ap)) %>%
  mutate(locality = case_when(
    o_ap >= -9.7 & o_ap <= -5.7 ~ "Local",
    TRUE ~ "Nonlocal"
  )) %>%
  mutate(locality = factor(locality, levels = c("Nonlocal", "Local")))

table(df_locality_subset$locality)
```

### Fit Bayesian model

```{r}
# δ13C (collagen)
model_c_col_loc <- brm(
  c_col ~ locality,
  data = df_locality_subset,
  prior = c(
    set_prior("normal(-8.21, 0.95)", class = "Intercept"),
    group_prior,
    common_priors
  ),
  iter = 4000, warmup = 1000, chains = 4, cores = 4, seed = 141
)

# δ15N (collagen)
model_n_air_loc <- brm(
  n_air ~ locality,
  data = df_locality_subset,
  prior = c(
    set_prior("normal(9.22, 1.02)", class = "Intercept"),
    group_prior,
    common_priors
  ),
  iter = 4000, warmup = 1000, chains = 4, cores = 4, seed = 142
)

# δ13C (apatite)
model_c_ap_loc <- brm(
  c_ap ~ locality,
  data = df_locality_subset,
  prior = c(
    set_prior("normal(-3.17, 0.69)", class = "Intercept"),
    group_prior,
    common_priors
  ),
  iter = 4000, warmup = 1000, chains = 4, cores = 4, seed = 143
)

# δ18O (apatite)
model_o_ap_loc <- brm(
  o_ap ~ locality,
  data = df_locality_subset,
  prior = c(
    set_prior("normal(-8.39, 1.32)", class = "Intercept"),
    group_prior,
    common_priors
  ),
  iter = 4000, warmup = 1000, chains = 4, cores = 4, seed = 144
)
```


### Posterior summaries

```{r}
summaries_loc <- list(
  tidy(model_c_col_loc, effects = "fixed", conf.int = TRUE) %>% mutate(model = "δ13C (collagen)"),
  tidy(model_n_air_loc, effects = "fixed", conf.int = TRUE) %>% mutate(model = "δ15N (collagen)"),
  tidy(model_c_ap_loc, effects = "fixed", conf.int = TRUE) %>% mutate(model = "δ13C (apatite)"),
  tidy(model_o_ap_loc, effects = "fixed", conf.int = TRUE) %>% mutate(model = "δ18O (apatite)")
)

posterior_table_loc <- bind_rows(summaries_loc) %>%
  select(model, term, estimate, std.error, conf.low, conf.high) %>%
  rename(
    Isotope_System = model,
    Parameter = term,
    Estimate = estimate,
    Std_Error = std.error,
    CI_2.5 = conf.low,
    CI_97.5 = conf.high
  )

print(posterior_table_loc)
```

### Posterior Probabilities: P(Local > Nonlocal)

```{r}
post_locality_probs <- tibble(
  Isotope_System = c("δ13C (collagen)", "δ15N (collagen)", "δ13C (apatite)", "δ18O (apatite)"),
  P_Local_gt_Nonlocal = c(
    mean(as_draws_df(model_c_col_loc)$b_localityLocal > 0),
    mean(as_draws_df(model_n_air_loc)$b_localityLocal > 0),
    mean(as_draws_df(model_c_ap_loc)$b_localityLocal > 0),
    mean(as_draws_df(model_o_ap_loc)$b_localityLocal > 0)
  )
)

print(post_locality_probs)
```

Due to sample size of Nonlocal data, variance comparison for locality cannot be run.



### Visualize data

```{r}
### Visualize data

ggplot(df_locality_subset, aes(x = c_col, y = locality, fill = locality)) +
  geom_density_ridges(alpha = 0.6, scale = 1, na.rm = TRUE) +
  labs(title = "δ13C (collagen) by Locality", x = "δ13C (‰)", y = "Locality") +
  theme_minimal() +
  theme(legend.position = "none")

ggplot(df_locality_subset, aes(x = n_air, y = locality, fill = locality)) +
  geom_density_ridges(alpha = 0.6, scale = 1, na.rm = TRUE) +
  labs(title = "δ15N (collagen) by Locality", x = "δ15N (‰)", y = "Locality") +
  theme_minimal() +
  theme(legend.position = "none")

ggplot(df_locality_subset, aes(x = c_ap, y = locality, fill = locality)) +
  geom_density_ridges(alpha = 0.6, scale = 1, na.rm = TRUE) +
  labs(title = "δ13C (apatite) by Locality", x = "δ13C (‰)", y = "Locality") +
  theme_minimal() +
  theme(legend.position = "none")

ggplot(df_locality_subset, aes(x = o_ap, y = locality, fill = locality)) +
  geom_density_ridges(alpha = 0.6, scale = 1, na.rm = TRUE) +
  labs(title = "δ18O (apatite) by Locality", x = "δ18O (‰)", y = "Locality") +
  theme_minimal() +
  theme(legend.position = "none")
```

---

## Diachronic comparisons (time periods)


## Subset data

```{r}
df_time_subset <- df_clean_isotopes %>%
  filter(time %in% c("Tlam", "Xol-Met", "Epi")) %>%
  mutate(
    time = recode(time, `Xol-Met` = "XolMet"),
    time = factor(time, levels = c("Tlam", "XolMet", "Epi"))
  ) %>%
  droplevels()

df_time_subset %>% count(time)

```

### Fit Bayesian model
```{r}
# d13C collagen
model_c_col_time <- brm(
  c_col ~ time,
  data = df_time_subset,
  prior = c(
    set_prior("normal(-8.21, 0.95)", class = "Intercept"),
    group_prior,
    common_priors
  ),
  iter = 4000, warmup = 1000, chains = 4, cores = 4, seed = 161,
  control = list(adapt_delta = 0.95)
)

# d15N
model_n_air_time <- brm(
  n_air ~ time,
  data = df_time_subset,
  prior = c(
    set_prior("normal(9.22, 1.02)", class = "Intercept"),
    group_prior,
    common_priors
  ),
  iter = 4000, warmup = 1000, chains = 4, cores = 4, seed = 162,
  control = list(adapt_delta = 0.95)
)

# d13C apatite
model_c_ap_time <- brm(
  c_ap ~ time,
  data = df_time_subset,
  prior = c(
    set_prior("normal(-3.17, 0.69)", class = "Intercept"),
    group_prior,
    common_priors
  ),
  iter = 4000, warmup = 1000, chains = 4, cores = 4, seed = 163,
  control = list(adapt_delta = 0.95)
)

# d18O apatite
model_o_ap_time <- brm(
  o_ap ~ time,
  data = df_time_subset,
  prior = c(
    set_prior("normal(-8.39, 1.32)", class = "Intercept"),
    group_prior,
    common_priors
  ),
  iter = 4000, warmup = 1000, chains = 4, cores = 4, seed = 164,
  control = list(adapt_delta = 0.95)
)
```

### Posterior summaries

```{r}
summaries_time <- list(
  tidy(model_c_col_time, effects = "fixed", conf.int = TRUE) %>% mutate(model = "δ13C (collagen)"),
  tidy(model_n_air_time, effects = "fixed", conf.int = TRUE) %>% mutate(model = "δ15N (collagen)"),
  tidy(model_c_ap_time, effects = "fixed", conf.int = TRUE) %>% mutate(model = "δ13C (apatite)"),
  tidy(model_o_ap_time, effects = "fixed", conf.int = TRUE) %>% mutate(model = "δ18O (apatite)")
)

posterior_table_time <- bind_rows(summaries_time) %>%
  select(model, term, estimate, std.error, conf.low, conf.high) %>%
  rename(
    Isotope_System = model,
    Parameter = term,
    Estimate = estimate,
    Std_Error = std.error,
    CI_2.5 = conf.low,
    CI_97.5 = conf.high
  )

print(posterior_table_time)
```

### Posterior Probabilities

```{r}
### Posterior Probabilities

# --- δ13C (collagen) ---
post_c_col_time <- as_draws_df(model_c_col_time)
p_c_col_xolmet_gt_tlam <- mean(post_c_col_time$b_timeXolMet > 0, na.rm = TRUE)
p_c_col_epi_gt_tlam    <- mean(post_c_col_time$b_timeEpi > 0, na.rm = TRUE)
p_c_col_epi_gt_xolmet  <- mean(post_c_col_time$b_timeEpi > post_c_col_time$b_timeXolMet, na.rm = TRUE)

# --- δ15N (collagen) ---
post_n_air_time <- as_draws_df(model_n_air_time)
p_n_air_xolmet_gt_tlam <- mean(post_n_air_time$b_timeXolMet > 0, na.rm = TRUE)
p_n_air_epi_gt_tlam    <- mean(post_n_air_time$b_timeEpi > 0, na.rm = TRUE)
p_n_air_epi_gt_xolmet  <- mean(post_n_air_time$b_timeEpi > post_n_air_time$b_timeXolMet, na.rm = TRUE)

# --- δ13C (apatite) ---
post_c_ap_time <- as_draws_df(model_c_ap_time)
p_c_ap_xolmet_gt_tlam <- mean(post_c_ap_time$b_timeXolMet > 0, na.rm = TRUE)
p_c_ap_epi_gt_tlam    <- mean(post_c_ap_time$b_timeEpi > 0, na.rm = TRUE)
p_c_ap_epi_gt_xolmet  <- mean(post_c_ap_time$b_timeEpi > post_c_ap_time$b_timeXolMet, na.rm = TRUE)

# --- δ18O (apatite) ---
post_o_ap_time <- as_draws_df(model_o_ap_time)
p_o_ap_xolmet_gt_tlam <- mean(post_o_ap_time$b_timeXolMet > 0, na.rm = TRUE)
p_o_ap_epi_gt_tlam    <- mean(post_o_ap_time$b_timeEpi > 0, na.rm = TRUE)
p_o_ap_epi_gt_xolmet  <- mean(post_o_ap_time$b_timeEpi > post_o_ap_time$b_timeXolMet, na.rm = TRUE)

# Combine into summary table
time_probs_summary <- tibble(
  Isotope_System = rep(c("δ13C (collagen)", "δ15N (collagen)", "δ13C (apatite)", "δ18O (apatite)"), each = 3),
  Comparison = rep(c("XolMet > Tlam", "Epi > Tlam", "Epi > XolMet"), 4),
  Probability = c(
    p_c_col_xolmet_gt_tlam, p_c_col_epi_gt_tlam, p_c_col_epi_gt_xolmet,
    p_n_air_xolmet_gt_tlam, p_n_air_epi_gt_tlam, p_n_air_epi_gt_xolmet,
    p_c_ap_xolmet_gt_tlam,  p_c_ap_epi_gt_tlam,  p_c_ap_epi_gt_xolmet,
    p_o_ap_xolmet_gt_tlam,  p_o_ap_epi_gt_tlam,  p_o_ap_epi_gt_xolmet
  )
)

print(time_probs_summary)
```

Variance models for time periods were explored but not reported due to small sample size in the Epi group, which produced unstable posterior estimates for group-specific standard deviations.


### Visualize Data
```{r}
# δ13C (collagen)
ggplot(df_time_subset, aes(x = c_col, y = time, fill = time)) +
  geom_density_ridges(alpha = 0.6, scale = 1, na.rm = TRUE) +
  scale_fill_manual(values = c("Tlam" = "#FFB347",
                               "XolMet" = "#87CEFA",
                               "Epi" = "#B0E57C")) +
  labs(title = "δ13C (collagen) by Time Period",
       x = "δ13C (‰)",
       y = "Time Period") +
  theme_minimal() +
  theme(legend.position = "none")

# δ15N (collagen)
ggplot(df_time_subset, aes(x = n_air, y = time, fill = time)) +
  geom_density_ridges(alpha = 0.6, scale = 1, na.rm = TRUE) +
  scale_fill_manual(values = c("Tlam" = "#FFB347",
                               "XolMet" = "#87CEFA",
                               "Epi" = "#B0E57C")) +
  labs(title = "δ15N (collagen) by Time Period",
       x = "δ15N (‰)",
       y = "Time Period") +
  theme_minimal() +
  theme(legend.position = "none")

# δ13C (apatite)
ggplot(df_time_subset, aes(x = c_ap, y = time, fill = time)) +
  geom_density_ridges(alpha = 0.6, scale = 1, na.rm = TRUE) +
  scale_fill_manual(values = c("Tlam" = "#FFB347",
                               "XolMet" = "#87CEFA",
                               "Epi" = "#B0E57C")) +
  labs(title = "δ13C (apatite) by Time Period",
       x = "δ13C (‰)",
       y = "Time Period") +
  theme_minimal() +
  theme(legend.position = "none")

# δ18O (apatite)
ggplot(df_time_subset, aes(x = o_ap, y = time, fill = time)) +
  geom_density_ridges(alpha = 0.6, scale = 1, na.rm = TRUE) +
  scale_fill_manual(values = c("Tlam" = "#FFB347",
                               "XolMet" = "#87CEFA",
                               "Epi" = "#B0E57C")) +
  labs(title = "δ18O (apatite) by Time Period",
       x = "δ18O (‰)",
       y = "Time Period") +
  theme_minimal() +
  theme(legend.position = "none")
```

---

## Visualization of posterior distributions

**Posterior Means**

**Compound**

```{r}
# δ13Ccol
means_c_col <- model_c_col %>%
  spread_draws(b_Intercept, b_frenteLV3) %>%
  mutate(
    LV2 = b_Intercept,
    LV3 = b_Intercept + b_frenteLV3
  ) %>%
  select(.draw, LV2, LV3) %>%
  pivot_longer(cols = c(LV2, LV3), names_to = "group", values_to = "value") %>%
  mutate(Isotope = "δ13Ccol", Type = "Mean")

# δ15N
means_n_air <- model_n_air %>%
  spread_draws(b_Intercept, b_frenteLV3) %>%
  mutate(
    LV2 = b_Intercept,
    LV3 = b_Intercept + b_frenteLV3
  ) %>%
  select(.draw, LV2, LV3) %>%
  pivot_longer(cols = c(LV2, LV3), names_to = "group", values_to = "value") %>%
  mutate(Isotope = "δ15N", Type = "Mean")

# δ13Cap
means_c_ap <- model_c_ap %>%
  spread_draws(b_Intercept, b_frenteLV3) %>%
  mutate(
    LV2 = b_Intercept,
    LV3 = b_Intercept + b_frenteLV3
  ) %>%
  select(.draw, LV2, LV3) %>%
  pivot_longer(cols = c(LV2, LV3), names_to = "group", values_to = "value") %>%
  mutate(Isotope = "δ13Cap", Type = "Mean")

# δ18O
means_o_ap <- model_o_ap %>%
  spread_draws(b_Intercept, b_frenteLV3) %>%
  mutate(
    LV2 = b_Intercept,
    LV3 = b_Intercept + b_frenteLV3
  ) %>%
  select(.draw, LV2, LV3) %>%
  pivot_longer(cols = c(LV2, LV3), names_to = "group", values_to = "value") %>%
  mutate(Isotope = "δ18O", Type = "Mean")

# Combine all means
all_means <- bind_rows(means_c_col, means_n_air, means_c_ap, means_o_ap)

all_means <- all_means %>%
  mutate(Isotope = case_when(
    Isotope == "δ13Ccol" ~ "δ13C (collagen)",
    Isotope == "δ13Cap"  ~ "δ13C (apatite)",
    Isotope == "δ15N"    ~ "δ15N (collagen)",
    Isotope == "δ18O"    ~ "δ18O (apatite)",
    TRUE ~ Isotope
  )) %>%
  mutate(Isotope = factor(Isotope, levels = c(
    "δ13C (collagen)", "δ13C (apatite)", "δ15N (collagen)", "δ18O (apatite)"
  )))

# Posterior mean lines for each group in each panel
mean_lines <- all_means %>%
  group_by(Isotope, group) %>%
  summarise(mean_value = mean(value, na.rm = TRUE), .groups = "drop")

# Panel labels with posterior probabilities
prob_labels <- tibble(
  Isotope = factor(c(
    "δ13C (collagen)",
    "δ13C (apatite)",
    "δ15N (collagen)",
    "δ18O (apatite)"
  ), levels = c(
    "δ13C (collagen)", "δ13C (apatite)", "δ15N (collagen)", "δ18O (apatite)"
  )),
  label = c(
    "P(LV2 > LV3) = 0.74",
    "P(LV2 > LV3) = 0.50",
    "P(LV2 > LV3) = 0.16",
    "P(LV2 > LV3) = 0.58"
  )
)

# Get annotation positions within each facet
label_positions <- all_means %>%
  group_by(Isotope) %>%
  summarise(
    x = quantile(value, 0.98, na.rm = TRUE),
    y = Inf,
    .groups = "drop"
  ) %>%
  left_join(prob_labels, by = "Isotope")

# Plot
p_means <- ggplot(all_means, aes(x = value, fill = group)) +
  geom_density(alpha = 0.6) +
  geom_vline(
    data = mean_lines,
    aes(xintercept = mean_value, color = group),
    linetype = "dashed",
    linewidth = 0.8,
    show.legend = FALSE
  ) +
  geom_text(
    data = label_positions,
    aes(x = x, y = y, label = label),
    inherit.aes = FALSE,
    hjust = 1.05,
    vjust = 1.5,
    size = 3.2
  ) +
  facet_wrap(~ Isotope, scales = "free", ncol = 2) +
  scale_fill_manual(values = c("LV2" = "darkorange", "LV3" = "steelblue")) +
  scale_color_manual(values = c("LV2" = "darkorange", "LV3" = "steelblue")) +
  labs(
    x = "Posterior Mean",
    y = "Density",
    fill = "Compound"
  ) +
  theme_minimal() +
  theme(
    strip.text = element_text(face = "bold"),
    legend.position = "right",
    text = element_text(size = 11)
  )

print(p_means)
```


**Sex**

```{r}

# δ13C collagen
means_c_col_sex <- model_c_col_sex %>%
  spread_draws(b_Intercept, b_sexM) %>%
  mutate(
    F = b_Intercept,
    M = b_Intercept + b_sexM
  ) %>%
  select(.draw, F, M) %>%
  pivot_longer(cols = c(F, M), names_to = "group", values_to = "value") %>%
  mutate(Isotope = "δ13Ccol")

# δ15N collagen
means_n_air_sex <- model_n_air_sex %>%
  spread_draws(b_Intercept, b_sexM) %>%
  mutate(
    F = b_Intercept,
    M = b_Intercept + b_sexM
  ) %>%
  select(.draw, F, M) %>%
  pivot_longer(cols = c(F, M), names_to = "group", values_to = "value") %>%
  mutate(Isotope = "δ15N")

# δ13C apatite
means_c_ap_sex <- model_c_ap_sex %>%
  spread_draws(b_Intercept, b_sexM) %>%
  mutate(
    F = b_Intercept,
    M = b_Intercept + b_sexM
  ) %>%
  select(.draw, F, M) %>%
  pivot_longer(cols = c(F, M), names_to = "group", values_to = "value") %>%
  mutate(Isotope = "δ13Cap")

# δ18O apatite
means_o_ap_sex <- model_o_ap_sex %>%
  spread_draws(b_Intercept, b_sexM) %>%
  mutate(
    F = b_Intercept,
    M = b_Intercept + b_sexM
  ) %>%
  select(.draw, F, M) %>%
  pivot_longer(cols = c(F, M), names_to = "group", values_to = "value") %>%
  mutate(Isotope = "δ18O")

# Combine
all_means_sex <- bind_rows(
  means_c_col_sex,
  means_n_air_sex,
  means_c_ap_sex,
  means_o_ap_sex
)

all_means_sex <- all_means_sex %>%
  mutate(Isotope = case_when(
    Isotope == "δ13Ccol" ~ "δ13C (collagen)",
    Isotope == "δ13Cap"  ~ "δ13C (apatite)",
    Isotope == "δ15N"    ~ "δ15N (collagen)",
    Isotope == "δ18O"    ~ "δ18O (apatite)",
    TRUE ~ Isotope
  )) %>%
  mutate(Isotope = factor(Isotope, levels = c(
    "δ13C (collagen)",
    "δ13C (apatite)",
    "δ15N (collagen)",
    "δ18O (apatite)"
  )))

# Posterior mean lines for each sex in each panel
sex_mean_lines <- all_means_sex %>%
  group_by(Isotope, group) %>%
  summarise(line_value = mean(value, na.rm = TRUE), .groups = "drop")

# Posterior probability labels
prob_labels_sex <- tibble(
  Isotope = factor(c(
    "δ13C (collagen)",
    "δ13C (apatite)",
    "δ15N (collagen)",
    "δ18O (apatite)"
  ), levels = levels(all_means_sex$Isotope)),
  label = c(
    "P(M > F) = 0.95",
    "P(M > F) = 0.68",
    "P(M > F) = 0.49",
    "P(M > F) = 0.78"
  )
)

# Plot
p_means_sex <- ggplot(all_means_sex, aes(x = value, fill = group)) +
  geom_density(alpha = 0.6) +
  geom_vline(
    data = sex_mean_lines,
    aes(xintercept = line_value, color = group),
    linetype = "dashed",
    linewidth = 0.8,
    show.legend = FALSE
  ) +
  geom_text(
    data = prob_labels_sex,
    aes(label = label),
    x = Inf,
    y = Inf,
    inherit.aes = FALSE,
    hjust = 1.1,
    vjust = 1.5,
    size = 4
  ) +
  facet_wrap(~ Isotope, scales = "free", ncol = 2) +
  scale_fill_manual(values = c("F" = "firebrick2", "M" = "deepskyblue")) +
  scale_color_manual(values = c("F" = "firebrick2", "M" = "deepskyblue")) +
  labs(
    x = "Posterior Mean",
    y = "Density",
    fill = "Sex"
  ) +
  theme_minimal() +
  theme(
    strip.text = element_text(face = "bold"),
    legend.position = "right",
    text = element_text(size = 11)
  )

print(p_means_sex)
```


**Locality**

```{r}
# δ13C collagen
means_c_col <- model_c_col_loc %>%
  spread_draws(b_Intercept, b_localityLocal) %>%
  mutate(
    Nonlocal = b_Intercept,
    Local = b_Intercept + b_localityLocal
  ) %>%
  select(.draw, Local, Nonlocal) %>%
  pivot_longer(cols = c(Local, Nonlocal), names_to = "group", values_to = "value") %>%
  mutate(Isotope = "δ13Ccol")

# δ15N collagen
means_n_air <- model_n_air_loc %>%
  spread_draws(b_Intercept, b_localityLocal) %>%
  mutate(
    Nonlocal = b_Intercept,
    Local = b_Intercept + b_localityLocal
  ) %>%
  select(.draw, Local, Nonlocal) %>%
  pivot_longer(cols = c(Local, Nonlocal), names_to = "group", values_to = "value") %>%
  mutate(Isotope = "δ15N")

# δ13C apatite
means_c_ap <- model_c_ap_loc %>%
  spread_draws(b_Intercept, b_localityLocal) %>%
  mutate(
    Nonlocal = b_Intercept,
    Local = b_Intercept + b_localityLocal
  ) %>%
  select(.draw, Local, Nonlocal) %>%
  pivot_longer(cols = c(Local, Nonlocal), names_to = "group", values_to = "value") %>%
  mutate(Isotope = "δ13Cap")

# δ18O apatite
means_o_ap <- model_o_ap_loc %>%
  spread_draws(b_Intercept, b_localityLocal) %>%
  mutate(
    Nonlocal = b_Intercept,
    Local = b_Intercept + b_localityLocal
  ) %>%
  select(.draw, Local, Nonlocal) %>%
  pivot_longer(cols = c(Local, Nonlocal), names_to = "group", values_to = "value") %>%
  mutate(Isotope = "δ18O")

# Combine
all_means_loc <- bind_rows(means_c_col, means_n_air, means_c_ap, means_o_ap)

all_means_loc <- all_means_loc %>%
  mutate(Isotope = case_when(
    Isotope == "δ13Ccol" ~ "δ13C (collagen)",
    Isotope == "δ13Cap"  ~ "δ13C (apatite)",
    Isotope == "δ15N"    ~ "δ15N (collagen)",
    Isotope == "δ18O"    ~ "δ18O (apatite)",
    TRUE ~ Isotope
  )) %>%
  mutate(Isotope = factor(Isotope, levels = c(
    "δ13C (collagen)", "δ13C (apatite)", "δ15N (collagen)", "δ18O (apatite)"
  )))

# Posterior mean lines for each locality group
loc_mean_lines <- all_means_loc %>%
  group_by(Isotope, group) %>%
  summarise(line_value = mean(value, na.rm = TRUE), .groups = "drop")

# Posterior probability labels
prob_labels_loc <- tibble(
  Isotope = factor(c(
    "δ13C (collagen)",
    "δ13C (apatite)",
    "δ15N (collagen)",
    "δ18O (apatite)"
  ), levels = levels(all_means_loc$Isotope)),
  label = c(
    "P(Local > Nonlocal) = 0.93",
    "P(Local > Nonlocal) = 0.99",
    "P(Local > Nonlocal) = 0.02",
    "P(Local > Nonlocal) = 1.00"
  )
)

# Plot
p_means_loc <- ggplot(all_means_loc, aes(x = value, fill = group)) +
  geom_density(alpha = 0.6) +
  geom_vline(
    data = loc_mean_lines,
    aes(xintercept = line_value, color = group),
    linetype = "dashed",
    linewidth = 0.8,
    show.legend = FALSE
  ) +
  geom_text(
    data = prob_labels_loc,
    aes(label = label),
    x = Inf,
    y = Inf,
    inherit.aes = FALSE,
    hjust = 1.1,
    vjust = 1.5,
    size = 4
  ) +
  facet_wrap(~ Isotope, scales = "free", ncol = 2) +
  scale_fill_manual(values = c("Local" = "springgreen2", "Nonlocal" = "purple3")) +
  scale_color_manual(values = c("Local" = "springgreen2", "Nonlocal" = "purple3")) +
  labs(
    x = "Posterior Mean",
    y = "Density",
    fill = "Locality"
  ) +
  theme_minimal() +
  theme(
    strip.text = element_text(face = "bold"),
    legend.position = "right",
    text = element_text(size = 11)
  )

print(p_means_loc)
```

**Time Period**

```{r}
# Extract posterior draws directly
extract_time_draws <- function(model, iso_label) {
  model %>%
    spread_draws(b_Intercept, b_timeXolMet, b_timeEpi) %>%
    mutate(
      Tlam = b_Intercept,
      `Xol-Met` = b_Intercept + b_timeXolMet,
      Epi = b_Intercept + b_timeEpi,
      Isotope = iso_label
    ) %>%
    select(.draw, Tlam, `Xol-Met`, Epi, Isotope) %>%
    pivot_longer(cols = c(Tlam, `Xol-Met`, Epi), names_to = "Time", values_to = "value")
}

# Extract draws
draws_c_col <- extract_time_draws(model_c_col_time, "δ13Ccol")
draws_n_air <- extract_time_draws(model_n_air_time, "δ15N")
draws_c_ap  <- extract_time_draws(model_c_ap_time, "δ13Cap")
draws_o_ap  <- extract_time_draws(model_o_ap_time, "δ18O")

# Combine into one dataframe
all_time_draws <- bind_rows(draws_c_col, draws_n_air, draws_c_ap, draws_o_ap)

all_time_draws <- all_time_draws %>%
  mutate(Isotope = case_when(
    Isotope == "δ13Ccol" ~ "δ13C (collagen)",
    Isotope == "δ13Cap"  ~ "δ13C (apatite)",
    Isotope == "δ15N"    ~ "δ15N (collagen)",
    Isotope == "δ18O"    ~ "δ18O (apatite)",
    TRUE ~ Isotope
  )) %>%
  mutate(Isotope = factor(Isotope, levels = c(
    "δ13C (collagen)", "δ13C (apatite)", "δ15N (collagen)", "δ18O (apatite)"
  )))

# Posterior mean lines for each time period
time_mean_lines <- all_time_draws %>%
  group_by(Isotope, Time) %>%
  summarise(line_value = mean(value, na.rm = TRUE), .groups = "drop")

# Probability labels for each facet
prob_labels_time <- tibble(
  Isotope = factor(c(
    "δ13C (collagen)",
    "δ13C (apatite)",
    "δ15N (collagen)",
    "δ18O (apatite)"
  ), levels = levels(all_time_draws$Isotope)),
  label = c(
    "P(Epi > Tlam) = 0.97",
    "P(Epi > Tlam) = 0.96",
    "P(Epi > Tlam) = 1.00",
    "P(Tlam > Epi) = 0.96"
  )
)

# Plot
p_time_means <- ggplot(all_time_draws, aes(x = value, fill = Time)) +
  geom_density(alpha = 0.6) +
  geom_vline(
    data = time_mean_lines,
    aes(xintercept = line_value, color = Time),
    linetype = "dashed",
    linewidth = 0.8,
    show.legend = FALSE
  ) +
  geom_text(
    data = prob_labels_time,
    aes(label = label),
    x = Inf,
    y = Inf,
    inherit.aes = FALSE,
    hjust = 1.1,
    vjust = 1.5,
    size = 3.5
  ) +
  facet_wrap(~ Isotope, scales = "free", ncol = 2) +
  scale_fill_manual(values = c("Tlam" = "#FFB347", "Xol-Met" = "#87CEFA", "Epi" = "#B0E57C")) +
  scale_color_manual(values = c("Tlam" = "#FFB347", "Xol-Met" = "#87CEFA", "Epi" = "#B0E57C")) +
  labs(
    x = "Posterior Mean",
    y = "Density",
    fill = "Time Period"
  ) +
  theme_minimal() +
  theme(
    strip.text = element_text(face = "bold"),
    legend.position = "right",
    text = element_text(size = 11)
  )

print(p_time_means)

```


**Posterior SD**

**Compound**

``` {r}
# δ13C (collagen)
sd_c_col <- model_c_col_sd %>%
  spread_draws(b_sigma_Intercept, b_sigma_frenteLV3) %>%
  mutate(
    LV2 = exp(b_sigma_Intercept),
    LV3 = exp(b_sigma_Intercept + b_sigma_frenteLV3)
  ) %>%
  select(.draw, LV2, LV3) %>%
  pivot_longer(cols = c(LV2, LV3), names_to = "group", values_to = "value") %>%
  mutate(Isotope = "δ13C (collagen)")

# δ13C (apatite)
sd_c_ap <- model_c_ap_sd %>%
  spread_draws(b_sigma_Intercept, b_sigma_frenteLV3) %>%
  mutate(
    LV2 = exp(b_sigma_Intercept),
    LV3 = exp(b_sigma_Intercept + b_sigma_frenteLV3)
  ) %>%
  select(.draw, LV2, LV3) %>%
  pivot_longer(cols = c(LV2, LV3), names_to = "group", values_to = "value") %>%
  mutate(Isotope = "δ13C (apatite)")

# δ15N (collagen)
sd_n_air <- model_n_air_sd %>%
  spread_draws(b_sigma_Intercept, b_sigma_frenteLV3) %>%
  mutate(
    LV2 = exp(b_sigma_Intercept),
    LV3 = exp(b_sigma_Intercept + b_sigma_frenteLV3)
  ) %>%
  select(.draw, LV2, LV3) %>%
  pivot_longer(cols = c(LV2, LV3), names_to = "group", values_to = "value") %>%
  mutate(Isotope = "δ15N (collagen)")

# δ18O (apatite)
sd_o_ap <- model_o_ap_sd %>%
  spread_draws(b_sigma_Intercept, b_sigma_frenteLV3) %>%
  mutate(
    LV2 = exp(b_sigma_Intercept),
    LV3 = exp(b_sigma_Intercept + b_sigma_frenteLV3)
  ) %>%
  select(.draw, LV2, LV3) %>%
  pivot_longer(cols = c(LV2, LV3), names_to = "group", values_to = "value") %>%
  mutate(Isotope = "δ18O (apatite)")

# Combine all isotopic SDs
all_sds <- bind_rows(sd_c_col, sd_c_ap, sd_n_air, sd_o_ap)

# Order facets
all_sds <- all_sds %>%
  mutate(Isotope = factor(Isotope, levels = c(
    "δ13C (collagen)",
    "δ13C (apatite)",
    "δ15N (collagen)",
    "δ18O (apatite)"
  )))

# Posterior mean SD values for dashed lines
sd_lines <- all_sds %>%
  group_by(Isotope, group) %>%
  summarise(line_value = mean(value, na.rm = TRUE), .groups = "drop")

# Posterior probability labels
prob_labels_sd <- tibble(
  Isotope = factor(c(
    "δ13C (collagen)",
    "δ13C (apatite)",
    "δ15N (collagen)",
    "δ18O (apatite)"
  ), levels = levels(all_sds$Isotope)),
  label = c(
    "P(LV2 > LV3) = 0.49",
    "P(LV2 > LV3) = 0.88",
    "P(LV2 > LV3) = 0.96",
    "P(LV2 > LV3) = 0.37"
  )
)

# Plot
p_sds <- ggplot(all_sds, aes(x = value, fill = group)) +
  geom_density(alpha = 0.6) +
  geom_vline(
    data = sd_lines,
    aes(xintercept = line_value, color = group),
    linetype = "dashed",
    linewidth = 0.8,
    show.legend = FALSE
  ) +
  geom_text(
    data = prob_labels_sd,
    aes(label = label),
    x = Inf,
    y = Inf,
    inherit.aes = FALSE,
    hjust = 1.1,
    vjust = 1.5,
    size = 4
  ) +
  facet_wrap(~ Isotope, scales = "free", ncol = 2) +
  scale_fill_manual(values = c("LV2" = "darkorange", "LV3" = "steelblue")) +
  scale_color_manual(values = c("LV2" = "darkorange", "LV3" = "steelblue")) +
  labs(
    x = "Posterior SD (σ)",
    y = "Density",
    fill = "Compound"
  ) +
  theme_minimal() +
  theme(
    strip.text = element_text(face = "bold"),
    legend.position = "right",
    text = element_text(size = 11)
  )

print(p_sds)
```

**Sex**

```{r}
# δ13C (collagen)
sd_c_col_sex <- model_c_col_sd_sex %>%
  spread_draws(b_sigma_Intercept, b_sigma_sexM) %>%
  mutate(
    F = exp(b_sigma_Intercept),
    M = exp(b_sigma_Intercept + b_sigma_sexM)
  ) %>%
  select(.draw, F, M) %>%
  pivot_longer(cols = c(F, M), names_to = "group", values_to = "value") %>%
  mutate(Isotope = "δ13C (collagen)")

# δ13C (apatite)
sd_c_ap_sex <- model_c_ap_sd_sex %>%
  spread_draws(b_sigma_Intercept, b_sigma_sexM) %>%
  mutate(
    F = exp(b_sigma_Intercept),
    M = exp(b_sigma_Intercept + b_sigma_sexM)
  ) %>%
  select(.draw, F, M) %>%
  pivot_longer(cols = c(F, M), names_to = "group", values_to = "value") %>%
  mutate(Isotope = "δ13C (apatite)")

# δ15N (collagen)
sd_n_air_sex <- model_n_air_sd_sex %>%
  spread_draws(b_sigma_Intercept, b_sigma_sexM) %>%
  mutate(
    F = exp(b_sigma_Intercept),
    M = exp(b_sigma_Intercept + b_sigma_sexM)
  ) %>%
  select(.draw, F, M) %>%
  pivot_longer(cols = c(F, M), names_to = "group", values_to = "value") %>%
  mutate(Isotope = "δ15N (collagen)")

# δ18O (apatite)
sd_o_ap_sex <- model_o_ap_sd_sex %>%
  spread_draws(b_sigma_Intercept, b_sigma_sexM) %>%
  mutate(
    F = exp(b_sigma_Intercept),
    M = exp(b_sigma_Intercept + b_sigma_sexM)
  ) %>%
  select(.draw, F, M) %>%
  pivot_longer(cols = c(F, M), names_to = "group", values_to = "value") %>%
  mutate(Isotope = "δ18O (apatite)")

# Combine all sex SDs
all_sds_sex <- bind_rows(
  sd_c_col_sex,
  sd_c_ap_sex,
  sd_n_air_sex,
  sd_o_ap_sex
)

# Set correct facet order
all_sds_sex <- all_sds_sex %>%
  mutate(Isotope = factor(Isotope, levels = c(
    "δ13C (collagen)",
    "δ13C (apatite)",
    "δ15N (collagen)",
    "δ18O (apatite)"
  )))

# Posterior mean SD lines for each sex in each panel
sex_sd_lines <- all_sds_sex %>%
  group_by(Isotope, group) %>%
  summarise(line_value = mean(value, na.rm = TRUE), .groups = "drop")

# Posterior probability labels
prob_labels_sd_sex <- tibble(
  Isotope = factor(c(
    "δ13C (collagen)",
    "δ13C (apatite)",
    "δ15N (collagen)",
    "δ18O (apatite)"
  ), levels = levels(all_sds_sex$Isotope)),
  label = c(
    "P(M > F) = 0.02",
    "P(M > F) = 0.02",
    "P(M > F) = 0.86",
    "P(M > F) = 0.32"
  )
)

# Plot
p_sds_sex <- ggplot(all_sds_sex, aes(x = value, fill = group)) +
  geom_density(alpha = 0.6) +
  geom_vline(
    data = sex_sd_lines,
    aes(xintercept = line_value, color = group),
    linetype = "dashed",
    linewidth = 0.8,
    show.legend = FALSE
  ) +
  geom_text(
    data = prob_labels_sd_sex,
    aes(label = label),
    x = Inf,
    y = Inf,
    inherit.aes = FALSE,
    hjust = 1.1,
    vjust = 1.5,
    size = 4
  ) +
  facet_wrap(~ Isotope, scales = "free", ncol = 2) +
  scale_fill_manual(values = c("F" = "firebrick2", "M" = "deepskyblue")) +
  scale_color_manual(values = c("F" = "firebrick2", "M" = "deepskyblue")) +
  labs(
    x = "Posterior SD (σ)",
    y = "Density",
    fill = "Sex"
  ) +
  theme_minimal() +
  theme(
    strip.text = element_text(face = "bold"),
    legend.position = "right",
    text = element_text(size = 11)
  )

print(p_sds_sex)
```
